Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-09 paper/main.tex
UPdates

Figure~\ref{fig:intro} highlights some of these issues. When constructing DAGs
manually, it is important to test how consistent our DAG is to our dataset. One
of the methods for checking the consistency is by testing whether the implied
CIs of the DAG hold in the dataset or not \citep{Ankan2021}. As each pair of
variables that do not have an edge between them in the DAG implies a CI
statement, we can use a statistical test to check whether that statement holds
in our dataset or not. If we find any violations we can also modify our DAG
based on this and utilizing our domain knowledge to find a suitable orientation
for the edge between the pair of variables.

@@

In this paper we present a method that utilizes this idea of CI testing to
assist researchers during the DAG construction instead of testing the final
model. We propose an iterative model construction method that combines domain
knowledge with data-driven insights. Our approach utilizes conditional
independence testing to identify pair of variables in the DAG that are missing
an edge or have a superfluous edge. Based on this information, in each
iteration we present one such pair of variables to an expert who then decides
the orientation of the edge between the variables based on their domain
knowledge. The method also iteratively prunes all superfluous edges in each
iteration. As a result this approach guides the expert in constructing their
model while ensuring that the model is consistent with the data.

As at any iteration, multiple changes to the DAG are possible. We also give a
way to rank potential new edges additions to the DAG using effect size measures
for CI tests.

>>>>

Figure~\ref{fig:intro} highlights some of these issues. On the other hand,
constructing DAG based solely on domain expertise can be error prone as it
requires us to distinguish between different causal structures, such as direct
or indirect effects or causal mediation. Given these complexities, it is
important to validate the consistency of a DAG against our dataset. One way to
test this consistency is by testing whether the CI statements implied by the
DAG hold in the data \citep{Ankan2021}. Specifically, each missing edge between
a pair of variables in the DAG implies a CI statement, which can be tested in
the dataset using a statistical test. If we find any violation to a CI
statement, we can fix it by adding an edge between the variables in the DAG. We
also need to decide the orientation of this edge based on domain expertise.

@@

In this paper, we propose a method that leverages this CI testing approach to
assist researchers during the DAG construction process, rather than just
validating the model. Our approach follows an iterative model-building strategy
that integrates domain expertise with CI testing on data. At each iterations,
we use CI testing to identify pairs of variables where an edge is either
missing or superfluous. These pairs are then presented to the expert, who
determines whether to add or remove these edges. The orientation of the edge
when adding is determined based on the domain expertise. As at any iteration,
multiple changes to the DAG are possible. We also give a way to rank potential
new edges additions to the DAG using effect size measures for CI tests to
address the most severe violations of CI tests first. As a result this approach
acts like a guide to the expert when manually constructing their model while
ensuring that the model is consistent with the data.
